A Tool for Visualizing Patterns of Spreadsheet Function Combinations

### 2016-04-21 doc/paper.tex
Progress from yesterday. Fixes #34, fixes #37, fixes #40, fixes #41, fixes #43, fixes #44, fixes #45

\begin{document} \title{A Tool for Visualizing Patterns of Spreadsheet Function
		Combinations}

	% author names and affiliations
	% use a multiple column layout for up to three different
	% affiliations
	\author{\IEEEauthorblockN{Justin A. Middleton} \IEEEauthorblockA{North Carolina
			State University\\ Raleigh, North Carolina}}

@@

	\section{Introduction} Business and research alike owe a lot to spreadsheets, the
	tabular interface that offers users a structured way to store and manipulate
	potentially huge bodies of data. Their allure is in their versatility: while
	the novice can still work without a deep knowledge of programming, they can
	learn to expedite their work by building programs out of several available
	functions \cite{nardi1990spreadsheet}. As such, it should not be surprising
	that when Scaffidi and colleagues estimated that over 50 million U.S. workers
	could be using them by 2012, with 25 million of them writing programs
	\cite{scaffidi2005estimating}. \par

	Considering this ubiquity, it's crucial to understand how people are actually
	using spreadsheets. After all, failures in their practice can be ruinous, as
	demonstrated in EuSpRIG's many horror stories of spreadsheets gone wrong. In
	2013, for example, they detail the influence of an economics paper on debt and
	national growth which shaped many political initiates around it, only to
	discover that the spreadsheets used in the data analysis contained an error
	that reversed the findings. The other stories, likewise, do not augur well if
	shoddy spreadsheet practices persist
	\footnote{http://www.eusprig.org/horror-stories.htm}. \par

	Fortunately, many spreadsheet researchers have assembled and released a number
	of spreadsheet corpora to inform work of how people actually use these tools.
	Collections like EUSES \cite{fisher2005euses}, FUSE \cite{barik2015fuse}, and
	the Enron corpus \cite{hermans2015enron} have already enabled fruitful work
	across the field. Hermans and colleagues, for example, used the EUSES corpus in
	2011 to inform and evaluate their work on exploring code smells, or inelegant
	places in code that suggest the need for refactoring, between worksheets
	\cite{hermans2012detecting}. Furthermore, these corpora have been compared
	and cross-referenced, such as in Jansen's contrast on Enron and EUSES, to
	gather insight in how different domains use spreadsheets differently 
	\cite{jansen2015enron}, or in Aivaloglou and colleague's evaluation of a
	spreadsheet grammar \cite{aivaloglou2015grammar}.\par

	A tool, then, that empowers users of any intention to explore these datasets
	would be rife with potential. On a simple level, the tool could provide
	concrete statistics on which functions are used in practice most (or least)
	often for a given dataset. Beyond this, though, it could also inform future
	work on recommending functions through an understanding of this patterns, or it
	could guide API improvement by suggesting functions to add or prune based how
	people actually use them. Furthermore, if the tool maintains the connection
	between its broad patterns and the actual instances of formulae in workbooks,
	it could serve as a boon to educators not only by showing which functions and
	combinations are most practical to learn but also by offering a bounty of
	instructive, and real, examples \par

	This paper contributes a tool, informed by some of the aforementioned
	spreadsheet corpora, that seeks to accomplish these tasks by offering an
	interactive interface for exploring how people combine functions. It seeks to
	balance several needs, such as capturing these patterns in a broad enough
	overview while still be specific enough lead back to the precise places in
	spreadsheets where these patterns are realized. Along with the tool, then, is
	also a discussion of the various goals which the tool seeks to fulfill and the
	trade-offs that we negotiated at each point. All of this culminates in an
	evaluation of the tool through a study of possible cases to which the tool
	might apply.

	\section{Related Work} This tool comes from a line of spreadsheet visualization
	tools before it, each with a different focus. Some tools, such as Igarashi and
	colleague's fluid visualizations \cite{igarashi1998fluid}, seek to visualize
	the hidden formulas within a spreadsheet by imposing the dataflow graphs over
	the cells. Likewise, Clermont \cite{clermont2003scalable} (and Hipfl
	\cite{hipfl2008using}, who extended his work) explored various ways of
	visualizing groups of related cells through similar functions, neighbors, or
	references. Others focus on creating visualizations outside of the sheets:
	Hermans and colleagues address a spreadsheet programmer's information needs
	through their work on the tools GyroSAT \cite{hermans2011supporting} and Breviz
	\cite{hermans2011breviz} to make dataflow diagrams of individual spreadsheets.
	These, however, tend to focus on individual spreadsheets and not bodies of
	them. \par

	Nevertheless, other studies and tools focus more on evaluating the content of
	the cells therein. Badame, for example, implements a number of Excel formula
	refactorings and uses Euses in part to evaluate it
	\cite{badame2012refactoring}. Hermans, leveraging her work with the
	aforementioned dataflow diagrams, approached the same dataset to assess the
	prevalence of common inter-worksheet smells, such as feature envy and
	inappropriate intimacy \cite{hermans2012detecting}. Cunha and colleagues,
	likewise, constructed and refined a catalogue of smells and implemented the
	SmellSheet Detective tool to detect bad smells in chosen spreadsheets
	\cite{cunha2012towards}, which Abreu and colleagues later extended to detect
	faults. \cite{abreu2014smelling}.\par

	Implicit in this study is an exploration of how users interact with the Excel
	API. Previous research, such as the first study on the Enron spreadsheet corpus
	\cite{hermans2015enron}, on how users employ certain functions or commands.
	Researchers have applied such questions to other contexts as well, such as
	Murphy and colleague's study into how Java developers use Eclipse and what
	commands they use most often \cite{murphy2006java}.  Others focus on the
	documentation of the API itself: Robillard and DeLine, for example conducted a
	series of surveys and interviews to diagnose common problems with learning
	APIs, and they found that code examples, one of the focuses of this tool, is
	one of the most important aids to have in learning a new system
	\cite{robillard2011field}.

@@

	\centerline{\includegraphics{start}}

>>>>

\usepackage{textcomp}
\newcommand{\toolname}{Perquimans?!}
\begin{document} \title{A Tool for Visualizing Patterns of Spreadsheet Function
		Combinations}

	% author names and affiliations
	% use a multiple column layout for up to three different
	% affiliations
	\author{
		\IEEEauthorblockN{Justin A. Middleton, Emerson Murphy-Hill} 
		\IEEEauthorblockA{North Carolina
			State University\\ Raleigh, North Carolina, USA}		
	}

@@

	\section{Introduction} Spreadsheets surround us, ordering the ever-growing
	flood of data that we produce. They serve key roles in industry, where as much
	as 95\% of U.S. financial firms use them daily \cite{panko2008sarbanes}, and academia,
	where teachers can use them to track research and students' grades, not to
	mention many other lines of work \cite{ko2011state}. It's this flexibility that
	grants them their allure: while a novice can still work without much experience,
	spreadsheets offer hundreds of specialized functions to fulfill a wide range of
	advanced needs, from complex statistics to text manipulation
	\cite{nardi1990spreadsheet}. As such, it should be no surprise that tens of
	millions of workers rely on these programs every day
	\cite{scaffidi2005estimating}.

	However, this ubiquity is not without danger: as more people adopt these tools
	and make larger and more complex sheets, the cost of failure grows too. To
	stress this, the European Spreadsheet Risks Interest Group (EuSpRIG) documents
	many horror stories of spreadsheets gone terribly
	wrong\footnote{http://www.eusprig.org/horror-stories.htm}. One such story tells
	of a 2013 economics paper which drew stark connections between debt and
	national growth. Its findings shaped many political initiatives throughout the
	United States and Europe. However, other researchers soon discovered that a
	critical coding error in the research spreadsheets hampered the results and
	that, by correcting this error, they completely reversed the findings! And this
	is only one of several accounts; even if most spreadsheet errors are relatively
	benign, poor spreadsheet practice can risk millions, if not billions, of
	dollars \cite{powell2009impact}.

	From these stories, we can see the dire importance of understanding how people
	actually use spreadsheets, if only to avoid disasters. Fortunately, spreadsheet
	researchers have supported this by amassing spreadsheet collections from
	diverse sources, industrial to academic. Collections like EUSES
	\cite{fisher2005euses}, Fuse \cite{barik2015fuse}, and the Enron corpus
	\cite{hermans2015enron} have already fueled productive research across the
	field, such as in Hermans' \cite{hermans2012detecting} or Jansen's
	\cite{jansen2015code} search for code smells\footnote{Or, instances of
		inelegant code that do not necessarily cause problems alone but nevertheless
		suggest a need for refactoring to prevent future problems. More on this in
		section IV.} in spreadsheet environments or Aivaloglou and colleague's work on
	making a grammar to capture every possible spreadsheet program in Excel
	\cite{aivaloglou2015grammar}. Additionally, some research has focused simply on
	comparing these collections, as Jansen did in 2015, to discover variation in
	spreadsheet techniques over time and location \cite{jansen2015enron}.

	Considering the sheer value of spreadsheets and all of these resources to describe
	their use, it becomes necessary that we design tools to make sense of it all. Tools,
	then, that empower people to explore datasets are rife with the potential to improve
	the current design of spreadsheets themselves and educate new generations of spreadsheets 
	users on how to work best.

	This paper contributes such a tool, \toolname, that attempts to fulfill this by building on the aforementioned
	spreadsheet corpora. Working with spreadsheets in Excel, \toolname focuses primarily
	on the functions, like SUM and IF and hundreds more, which are the building blocks
	of formula and programs which determine values across a spreadsheet. The tool creates
	an interactive, exploratory visualization which captures how people combine these functions
	in a selected collection of spreadsheets, quantifying the broad patterns of function nesting
	while also being specific enough to find anomalous formulae and link them
	back to the specific cell in spreadsheets where they originate. \par

	Alongside the tool, we discuss several of the key decisions we faced during design,
	such as what to do with redundant formulae in a sheet or how to represent optional
	arguments, and we describe how our overall design goals both helped and hindered this
	process. We then apply the tool to a number of case studies to evaluate how much it
	could contribute to several different contexts:

	\begin{itemize}
		\item API improvement
		\item Detection of code smells
		\item Spreadsheet education
	\end{itemize}

	To round out that discussion, we also conducted a user study to judge how usable 
	and intuitive the tool actually is. This grounds the discussion of its potential uses
	while also underscoring some of its inherent limitations and setting the way
	for future extensions and improvements upon the concept.

	\section{Related Work} Given humanity's affinity for visual information,
	visualizations have long been prized for their ability to communicate
	information quickly and efficient \cite{baeza1999modern}; the visualization of
	spreadsheets, therefore, has offered much room for research. Several previous
	tools work within a spreadsheet, seeking to clarify the connections between the
	visible values in cells, the formulae that produce them, and the dependencies
	between cells that inform them. Igarashi and colleague's fluid visualizations
	\cite{igarashi1998fluid}, for example, do this by imposing lines, color, and
	animation over the spreadsheet to visually explain these tangled connections.
	Likewise, Clermont explored ways of grouping cells by color and border that
	were related through similar functions, neighbors, or references
	\cite{clermont2003scalable}, and Hipfl extended this work by incorporating the
	layouts and labels of spreadsheets \cite{hipfl2008using}. Other tools focus on
	creating visualizations external to the sheets. Hermans and colleagues address a
	spreadsheet programmer's information by making dataflow diagrams from individual sheets
	through the tools GyroSAT \cite{hermans2011supporting} and Breviz \cite{hermans2011breviz}. 
	\toolname is among the latter class, creating standalone visualizations about spreadsheets. In contrast to Hermans' work, though, our work focuses on representing
	functions and how they comprise complex formula, rather than exploring the data
	structures of any particular spreadsheet.  \par

	Many other tools inspect the structure of formulas but often from the
	perspective of finding bad design.  Following Fowler's initial description of
	code smells in object-oriented design \cite{fowler2009refactoring}, many
	researchers have applied similar observations to spreadsheets and have
	cataloged classes of problematic formulae \cite{hermans2012detecting}
	\cite{cunha2012towards} \cite{asavametha2012detecting}, often accompanying them
	with detection tools \cite{abreu2014smelling} and investigations \cite{jansen2015code}. For example, Hermans, leveraging her work with the aforementioned dataflow diagrams, has worked with assessing common inter-worksheet smells, such as feature envy and
	inappropriate intimacy, wherein references between worksheets (rather
	than cells) in a file suggest a problem. \cite{hermans2012detectinginter}. Other
	tools, like Badam's Excel refactoring \cite{badame2012refactoring}, are concerned
	more with understanding the formula in the service of changing them. 	
	However, these studies define standards of poor design and orient their searches
	around them. Though \toolname supports the search for bad design, as we will see, 
	the tool itself takes a much more agnostic approach in presentation, offering 
	examples without regard for design quality. \par

	Implicit in this study is an exploration of the Excel API and how people use
	it. Previous research, such as the first study on the Enron spreadsheets
	\cite{hermans2015enron} or in its comparison with the EUSUS corpus
	\cite{jansen2015enron}, quantify how ofter users employ certain functions but
	not how they combine them. Researchers have applied such questions to other
	contexts as well, such as Murphy and colleague's study into how Java developers
	use Eclipse and what commands they use most often \cite{murphy2006java}. 
	Others focus on the documentation of the API itself: Robillard and DeLine, for
	example conducted a series of surveys and interviews to diagnose common
	problems with learning APIs, and they found that code examples, one of the
	focuses of this tool, is one of the most important aids to have in learning a
	new system \cite{robillard2011field}. Though this paper applies the tool to the
	Enron corpus specifically, the tool itself is not limited to this; it can be 
	readily applied to any collection of Excel spreadsheets a user gives it.

@@

	\begin{figure}
		\centering
		\includegraphics{start}
		\label{fig:startpic}
		\caption{How the IF tree looks at the start}
	\end{figure}
